---
title: "ACIDD_TOC"
author: "Nicholas Baetge"
date: "8/28/2020"
output: github_document
---

# Intro

This document shows how **individual bottle** DOC data from ACIDD remineralization bioassays were processed, QC'd, and analyzed. 

```{r message = F, warning = F}
library(tidyverse) 
library(readxl)
library(lubridate)
```

# Import data

Just like the bacterial abundance data, the DOC data exists in two separate sheets of an excel file. We'll take a look at what is in those sheets and combine the data using a "join" function

```{r}
excel_sheets("~/GITHUB/eemb144l/Input_Data/week4/ACIDD_Exp_TOC.xlsx") #let's see what the excel sheets are called

metadata <- read_excel("~/GITHUB/eemb144l/Input_Data/week4/ACIDD_Exp_TOC.xlsx", sheet = "Metadata") #import the first sheet containing metadata
names(metadata)

data <- read_excel("~/GITHUB/eemb144l/Input_Data/week4/ACIDD_Exp_TOC.xlsx", sheet = "Data") #import the second sheet containing the cell abundances
names(data)

joined <- left_join(metadata, data)
names(joined)
summary(joined)
View(joined)

```

# Prepare data

Let's convert the date and time column values from characters to dates, add a column with the time elasped for each experiment (in days), and then rearrange the columns. 

```{r warning = F}
toc <- joined %>% 
  mutate(Datetime = ymd_hm(Datetime)) %>% 
  group_by(Experiment, Bottle) %>% 
  mutate(interv = interval(first(Datetime), Datetime), 
         hours = interv/3600,
         days = hours/24) %>% 
  ungroup() %>% 
  rename(sd_TOC = TOC_sd,
         sd_PTOC = PTOC_sd) %>% 
  select(Experiment:Datetime, hours, days, everything(), -interv)
``` 

# Plot the curves

We have multiple organic carbon measurements: TOC and P(arallel) TOC. We'll plot all of them. We could plot each parameter separately, but one thing we can do to streamline plotting is to first "lengthen" our data (increase rows, decrease columns) using pivot_longer().

**We only have data for the Santa Barbara experiments, so we'll filter out the San Diego experiments for now**

## Pivot data

```{r}
pivot_toc <- toc %>% 
  select(Experiment, Location, Bottle, Treatment, days, TOC, PTOC ) %>% 
  pivot_longer(TOC:PTOC, names_to = "sample", values_to = "value")

pivot_toc_sd <- toc %>% 
  select(Experiment, Location, Bottle, Treatment, days, sd_TOC, sd_PTOC) %>% 
  pivot_longer(sd_TOC:sd_PTOC, names_to = "sample", names_prefix = "sd_", values_to = "sd")

pivoted <- left_join(pivot_toc, pivot_toc_sd) %>% 
  mutate(sample = ifelse(sample == "TOC", "Bottle", "Vial"))

View(pivoted)

```


## Plot

```{r  fig.height = 4, fig.width = 10, fig.align = "center"}

custom.colors <- c("Control" = "#377EB8", "Ash Leachate" = "#4DAF4A", "Santa Barbara" = "#E41A1C", "San Diego" = "#FF7F00", "Vial" = "#377EB8", "Bottle" = "#E41A1C")
levels <- c("Control", "Ash Leachate", "San Diego", "Santa Barbara", "Bottle", "Vial")

pivoted %>% 
  filter(Location == "Santa Barbara") %>% 
  drop_na(value) %>% 
  ggplot(aes(x = days, y = value, group = interaction(Experiment, Treatment, Bottle)) ) + 
  geom_errorbar(aes(ymin = value - sd, ymax = value + sd, color = factor(sample, levels = levels)), width = 0.4) +
  geom_point(aes(fill = factor(sample, levels = levels)), size = 3, shape = 21, alpha = 0.7 ) +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  labs(x = "Days", y = expression(paste("Total Organic Carbon, µmol C L"^-1))) +
  theme_classic() +
  facet_grid(~factor(Treatment, levels = levels), scales = "free") +
  theme(legend.title = element_blank()) +
  guides(color = F)
```


# TOC v PTOC

Let's do a simple regression comparing TOC and PTOC samples. There will be only 4 points here, but this may be a useful exercise for your data. 

```{r echo = FALSE}
library(lmodel2)

reg.data <- toc %>% 
  drop_na(TOC) %>% 
  filter(Timepoint > 0)

reg <- lmodel2(PTOC ~ TOC, data = reg.data, nperm = 99)
```

```{r echo = FALSE}
reg
```


```{r fig.height = 4, fig.width = 5, fig.align = "center"}
reg.data %>% 
  ggplot(aes(x = TOC, y = PTOC)) + 
  geom_abline(intercept = reg$regression.results[3,2],
              slope = reg$regression.results[3,3],colour = "black", linetype = 2, size = 1) +
   geom_abline(intercept = reg$confidence.intervals[3,2],
              slope = reg$confidence.intervals[3,4],colour = "grey", linetype = 3, size = 1) +
  geom_abline(intercept = reg$confidence.intervals[3,3],
              slope = reg$confidence.intervals[3,5],colour = "grey", linetype = 3, size = 1) +
  geom_errorbar(aes(ymin = PTOC - sd_PTOC, ymax = PTOC + sd_PTOC), width = 0.05) +
  #geom_errorbarh(aes(xmin = TOC - sd_TOC, xmax = TOC + sd_TOC), height = 0.01) +
  geom_point( shape = 21, color = "black", size = 4, alpha = 1) +
  labs(x = expression(paste("Bottle TOC, µmol C L"^-1)), y = expression(paste("Vial TOC"^-1))) +
  theme_classic() +
  annotate( geom = "text", label = expression(atop("y = 0.72x + 19.8", paste("r"^2," = 0.97, ", italic("p "), "= 0.01"))), x = 70, y = 74, size = 4) 
```

The vial and bottle TOC samples are not significantly different from one another and in fact, the difference in the measured values are within analytical detection limit for our method (1.5 µmol C L^-1^).

We can move forward with our analysis using the vial TOC values, since the dataset for those samples are more complete.

# Merge bacterial carbon data

We're particulary interested in DOC, the substrate used by bacterioplankton. We can calculate DOC as the difference between TOC and bacterial carbon. To do so, we'll need to merge this dataset with the processed bacterial abundance dataset. 

```{r}
bc <- read_rds("~/GITHUB/eemb144l/Output_Data/week3/ACIDD_Exp_Processed_BactAbund.rds")

merge <- full_join(bc, toc) %>% 
  arrange(Experiment, Bottle, Timepoint) %>% 
  fill(exp_start:delta_cells, delta_bc:sd_lag) #fill the values for variables that are the same for every timepoint of every treatment
  
```

Ok, so we don't have a TOC sample for every timepoint in which we sampled bacterial abundance and vice-versa. Before we move forward, let's first linearly interpolate the TOC and bacterial carbon data.  

# Interpolate

To interpolate NA observations, we'll use the zoo() and na.approx() functions in the package zoo. 

```{r}
library(zoo)
```

We only want to interpolate within an experiment and not across experiments, so we'll perform the interpolation by using the split-apply-combine strategy. We'll:

- split the dataframe into a list of its elements (in this case, experiments and bottles)
- apply the interpolation function to each of the list elements 
- combine the results into a new dataframe

## Split

```{r warning = FALSE}

to_interpolate <- merge %>%  
  select(Experiment, Bottle, Timepoint, days, PTOC, bc) %>% 
  group_by(Experiment, Bottle)  # first we'll define the grouping of our dataframe

list <- to_interpolate %>%
  group_split() #then we can convert the dataframe into a list, broken up by the groups (list elements)

#notice that the list elements don't have names. we can add those ourselves
keys <- to_interpolate %>% 
  group_keys() %>%
  mutate(key = paste(Experiment, Bottle))
header <- keys$key

names(list) <- header

```

## Write the function

```{r}
interp.func <- function(x) {
  y <- zoo(x, order.by = x$days) #orders our observations
  interp_toc <- round(as.numeric(na.approx(y$PTOC, na.rm = F)), 1) #interpolate doc
  interp_bc <- round(as.numeric(na.approx(y$bc, na.rm = F)), 1) #interpolate bc
  z <- cbind(y, interp_toc, interp_bc) #combine the columns 
  as_tibble(z) #convert to dataframe
}
```

## Apply and Combine

```{r}

interpolated <- lapply(list, interp.func) %>% 
  plyr::ldply(., as.data.frame) %>% 
  select(-c(.id, PTOC, bc, days)) %>% 
  mutate_at(vars(Timepoint:interp_bc), as.numeric) %>% 
  left_join(merge, .)
```

# Estimate DOC, Bioavailability, BGEs

```{r}
doc <- interpolated %>% 
  mutate(doc = interp_toc - interp_bc) %>% 
  group_by(Experiment, Bottle) %>%
  mutate(bioav_doc = (first(doc) - last(na.omit(doc))) / first(doc),
         doc_exp_end = ifelse(Timepoint == exp_end, doc, NA),
         delta_doc = first(doc) - doc_exp_end,
         bge = ifelse(delta_doc > 1.5, delta_bc/delta_doc, NA)) %>% 
  fill(doc_exp_end:bge, .direction = "downup") %>% 
  ungroup()
  
```


# Calculate treatment averages

```{r}
averages <- doc %>%
  group_by(Experiment, Treatment, Timepoint) %>% 
  mutate(ave_toc = mean(PTOC),
         sd_toc = sd(PTOC)) %>% 
  ungroup() %>% 
  group_by(Experiment, Treatment) %>% 
  mutate(ave_bioav_doc = mean(bioav_doc),
         sd_bioav_doc = sd(bioav_doc),
         ave_delta_doc = mean(delta_doc),
         sd_delta_doc = sd(delta_doc),
         ave_bge = mean(bge),
         sd_bge = sd(bge)) %>% 
  ungroup()
  
```

# Plot treament curves

```{r}
averages %>% 
  filter(Location == "Santa Barbara") %>% 
  drop_na(ave_toc) %>% 
  ggplot(aes(x = days, y = ave_toc, group = interaction(Experiment, Treatment)) ) + 
  geom_errorbar(aes(ymin = ave_toc - sd_toc, ymax = ave_toc + sd_toc, color = factor(Treatment, levels = levels)), width = 0.4) +
  geom_point(aes(fill = factor(Treatment, levels = levels)), size = 3, shape = 21, alpha = 0.7 ) +
   geom_line(aes(color = factor(Treatment, levels = levels)), size = 1, alpha = 0.7 ) +
  scale_color_manual(values = custom.colors) +
  scale_fill_manual(values = custom.colors) +
  labs(x = "Days", y = expression(paste("Total Organic Carbon, µmol C L"^-1))) +
  theme_classic() +
  facet_grid(~factor(Location, levels = levels), scales = "free") +
  theme(legend.title = element_blank()) +
  guides(color = F)
```

# Plot treatment averages 

```{r}
bar.data <- averages %>% 
  select(Location, Treatment, ave_bioav_doc:sd_bge) %>% 
  distinct() %>% 
  filter(Location == "Santa Barbara")
  
```


```{r}
bioav <- bar.data %>% 
ggplot(aes(x = factor(Treatment, levels = levels), y =  ave_bioav_doc, group = interaction(Location, Treatment)))  + 
  geom_bar(position = position_dodge(), stat = "identity", color = "black", fill = "white", alpha = 1) +
  geom_errorbar(aes(ymin = ave_bioav_doc - sd_bioav_doc, ymax = ave_bioav_doc + sd_bioav_doc), position = position_dodge(width = 0.9), stat = "identity", width = 0.1, size = 0.5) +
  labs(x = "", y = expression(paste("Bioavailable DOC Fraction")), color = "") +
  theme_classic() +
  #facet_grid(~factor(Location, levels = levels), scales = "free") +
  guides(fill = F) 

bioav
```



```{r}
deldoc <- bar.data %>% 
ggplot(aes(x = factor(Treatment, levels = levels), y =  ave_delta_doc, group = interaction(Location, Treatment)))  + 
  geom_bar(position = position_dodge(), stat = "identity", color = "black", fill = "white", alpha = 1) +
  geom_errorbar(aes(ymin = ave_delta_doc - sd_delta_doc, ymax = ave_delta_doc + sd_delta_doc), position = position_dodge(width = 0.9), stat = "identity", width = 0.1, size = 0.5) +
  labs(x = "", y = expression(paste("∆ DOC, µmol C L"^-1)), color = "") +
  theme_classic() +
  #facet_grid(~factor(Location, levels = levels), scales = "free") +
  guides(fill = F) 

```


```{r}
bge <- bar.data %>% 
ggplot(aes(x = factor(Treatment, levels = levels), y =  ave_bge, group = interaction(Location, Treatment)))  + 
  geom_bar(position = position_dodge(), stat = "identity", color = "black", fill = "white", alpha = 1) +
  geom_errorbar(aes(ymin = ave_bge - sd_bge, ymax = ave_bge + sd_bge), position = position_dodge(width = 0.9), stat = "identity", width = 0.1, size = 0.5) +
  labs(x = "", y = expression(paste("Bacterial Growth Efficiency")), color = "") +
  theme_classic() +
  #facet_grid(~factor(Location, levels = levels), scales = "free") +
  guides(fill = F) 

```


```{r}
library(patchwork)
```

```{r fig.width = 4, fig.height = 8}
deldoc / bioav / bge  + plot_annotation(title = "Santa Barbara", tag_levels = "a")
```

# Save Data

```{r}
saveRDS(averages, "~/GITHUB/eemb144l/Output_Data/week4/ACIDD_Exp_Processed_DOC_BGE.rds")
```





